\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2017-01-24 doc/publications/plos16/plos16.tex
Mostly typo corrections

\section{Introduction}
The digital revolution is seeing a dramatic increase in data collected about almost every aspect of
life~\cite{Loebbecke2015}.  These data sets are not only growing vertically, by capturing more events, but also
horizontally by capturing more information about these events.  The challenge of big and ``wide'' data is especially
pronounced in the health space where, for example, whole genome sequence (WGS) technology enables researchers to
interrogate all 3 billion base pairs of the human genome.

@@

For classifiction problems there is no obvious decomposition of classification error into an equivalent bias plus
variance form. A number of different additive decompositions of classification error have been considered in an effort
to understand the success of bagging. They all less transparent than the regression case (see \cite{Friedman.1997} and
citations therein).

@@

Since the original implementation of random forests, there have been a number of developments leading to a better
understanding of the algorithm.
\begin{itemize}
\item random forests are resistant to overfitting. However it is not true that they will not overfit,
  ~\cite{Segal.2004} gives examples of overfitting in cases where:
  \begin{itemize}
  \item a single tree is the most appropriate model; and
  \item the variables are highly correlated so that the bootstrap resampling does not give rise to uncorrelated trees.
  \end{itemize}
However, keeping these difficult cases in mind, random forests have a strong claim to being resistant to overfitting.
\item \cite{Strobl.et.al.2007} demonstrates that the random forest algorithm is subject to a bias in variable selection
  via two mechanisms:
  \begin{enumerate}
  \item differences in variable scale (for $X_i$ continuous) and number of levels (for $X_i$ discrete) introduces a
    bias. A uninformative variable with a large number of levels may be selected over a more informative variable with
    fewer levels;
  \item bagging introduces a bias, but sampling $(1- 1/e) \approx 0.632$ of the data without replacement seems to be an
    effective strategy for removing the problem.
  \end{enumerate}
\end{itemize}

@@

Biau {\it et al.}~\cite{Biau.2012} provide a proof that a random forest model will converge at a rate that depends on
the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
rate that depends on the size of $S$ and not the number of noise variables in the data set. If the size of $S$ is small
compared to the total number of variables, then the rate of convergence will be quicker. It is not clear what the size
of the set $S$ will be in a GWAs application. There are cases of small groups of SNPs with large effects and also of
conditions caused by the interaction of large groups of SNPs.

@@

However, within an RDD, spark ML stores each sample as a vector. Unlike RDDs, which can be partitioned and distributed
across multiple nodes, each vector must be present in its entirity on any node accessing it. This is no problem with
typical datasets, however as dimensionality increases, the vectors eventually reach a size where they can no longer fit
into a single node's memory.

@@

The current implementation of \cursedforest uses an ``Gini impurity'' criteria for splitting. Let $f_q$ be the fraction
of items labelled with value $q$ where $q = 1, \ldots, Q$ at a node. The Gini impurity is
$$
I_G(f) = \sum_{q = 1}^Q f_q ( 1 - f_q ), 
$$
which is a minimum when all observatkions at the node are in the same class.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This section talks about the datasets
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Datasets}
\subsubsection{1000 Genomes Project}
We obtained the 1000 Genomes Project data as VCF files from their FTP site.  Each VCF file contains the variants for
every individual across one chromosome.  We use the phase 3 dataset which contains 2,504 individuals with approximately
81 million variant positions (i.e. a matrix of $2,504 \times 81$ million).  For compatibility with the majority of the
algorithms we demonstrate, we convert the original variant representations to double representations.  For example, we
store the absence of a variant (0\textbar0) as 0, a heterozygous
variant (1\textbar0 or 0\textbar1) as 1, and a
homozygous variant (1\textbar1) as 2.  This is not the only possible encoding.  \cite{Goldstein.et.al.2011} note that of
the encoding listed in Table~\ref{table:encodings}, the additive encoding will present the most difficulty for a random
forest model.

@@

\subsection{Parameter settings}
We consider the parameter setting for the random forest algorithm. We use the R notation from the \randomforest
package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
the advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree{}  -- the number of trees.  The number of trees necessary for good performance grows with the number of
  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree is necessary to get stable estimates of variable
  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
  it is possible for a random forest model to have a poorer fit and still have an accurate ranking of variable
  importance;
\item \mtry{}  -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry may give
  better performance;
\item the size and complexity of the individual trees is controlled in \randomforest by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest\footnote{The Spark ML documentation \cite[]{Spark.2016} sets \texttt{maxDepth} to 4 in their
    classification example. This may give an estimate with a higher bias and, as such, may be a poor choice. See the
    discussion in \cite{Dietterich.2002} where it is shown that bagging may reduce bias as well as variance but gives a
    better result for low bias learners. \cite{Dietterich.2002} notes that if the boostrap replicate approximation were
    correct (i.e.~if the bootstrap sample came from an identical distribution to the data), then bagging would reduce
    variance without changing bias.} In the TCGA example the RandomForests code produced trees with depths of 30+.
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 2: Feature Selection on real data 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ selects previously demonstrated bone-density variants}
We investigate the ability of using a random forest model we build on genomic data to identify variants related to a
specific phenotype.
In this case we aim to identify variants associated with Bone Minderal Density (BMD). We build a model on the BMD dataset
(see datasets) using a binary high or low BMD value as the label.
From this model, we output a list of the top 10,000 contributing variants, ranked by their Gini impurity. The variant with
the highest Gini impurity---and hence, importance---is ranked 1, and the variant with the lowest Gini impurity is ranked 10,000.
Then, using the 26 known and 6 novel BMD genes from Duncan et al., we compile a list of variants that are present in or near one of these
32 genes.

@@

We also frequently see highly-ranked variants in chromosome 22q11. Stagi et al. observe that individuals with a microdeletion
in this region have a significant in bone mass\cite{Stagi2010}. In this same region, we also observe variants present in
PRODH (proline dehydrogenase). PRODH catalyzes the first step in proline degredation, and as proposed by Phang et al., proline
is made available by collagen degradation \cite{Phang2008}.

DCDC5 (Doublecortin Domain Containing 5) is one of the 26 known genes, however it is one of the genes that Duncan et al.
failed to replicate in their study. DCDC5 does appear in our list of genes, however the first occurance of a variant in
or near this gene has a rank of 9,667 out of 10,000. Also, this gene is not highly expressed in bone, and any mechanisms
by which this gene might regulate BMD are unclear \cite{Thakker2012}. Although variants near DCDC5 are associated with
BMD \cite{Rivadeneira2009}, it is questionable as to whether the gene itself contributes to BMD.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 3: Classification selection
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest outperforms existing methods for classification problems}
In this section we compare \cursedforest 's performance against other published methods for predicting ethnicity on the
1000 Genomes Project dataset.

@@

\begin{figure}[tbhp] 
    \caption{\textbf{Runtime comparison on the 1000 genomes project}} 
    \label{figure:synth1}
      \includegraphics[totalheight=8cm]{./figs/1000genomesRuntime.pdf}
    \begin{flushleft} 
      The figure shows the runtime for the 5 subsets of the 1000 genomes project and the different methods investigated. $\bigotimes$ marks the last dataset the different method completed before running out of memory or running more than 24 hours. The dashed lines mark the trajectory for how ranger and SparkML had performed given larger memory machines. 
    \end{flushleft}
\end{figure}

@@

We note several things here:
\begin{itemize}
\item the Donoho-Tanner phase transition arises in recovering the $\beta$ in data generated by a
  linear model. However, in a decision tree (random forest) there is no notion of estimating the $\beta$
\item A decision tree (random forest) is a heruistic search. It may recover a relationship in the space of the
  combinatorial search.
\end{itemize}

The existence of the  Donoho-Tanner phase transition is a salutory warning. There are likley to be limits, both
computionaal and logical, to the recovery of signals from noisy data. \cursedforest  is a contribution to addressing the
practivcal limits but the logical limits will still apply. However in the case of data that is both big and wide,
\cursedforest and other VariantSpark methods may provide a useful tool.

>>>>

\section{Introduction}
The digital revolution is seeing a dramatic increase in data collected about almost every aspect of
life~\cite{Loebbecke2015}.  These data sets are not only growing vertically, by capturing more events, but also
horizontally by capturing more information about these events.  The challenge of big and ``wide'' data is especially
pronounced in the health space where, for example, whole genome sequencing (WGS) technology enables researchers to
interrogate all 3 billion base pairs of the human genome.

@@

For classifiction problems there is no obvious decomposition of classification error into an equivalent bias plus
variance form. A number of different additive decompositions of classification error have been considered in an effort
to understand the success of bagging. They are all less transparent than the regression case (see \cite{Friedman.1997} and
citations therein).

@@

Since the original implementation of random forests, there have been a number of developments leading to a better
understanding of the algorithm.
\begin{itemize}
\item random forests are resistant to overfitting. However it is not true that they will not overfit,
  ~\cite{Segal.2004} gives examples of overfitting in cases where:
  \begin{itemize}
  \item a single tree is the most appropriate model; and
  \item the variables are highly correlated so that the bootstrap resampling does not give rise to uncorrelated trees.
  \end{itemize}
However, keeping these difficult cases in mind, random forests have a strong claim to being resistant to overfitting.
\item \cite{Strobl.et.al.2007} demonstrates that the random forest algorithm is subject to a bias in variable selection
  via two mechanisms:
  \begin{enumerate}
  \item differences in variable scale (for $X_i$ continuous) and number of levels (for $X_i$ discrete) introduces a
    bias. An uninformative variable with a large number of levels may be selected over a more informative variable with
    fewer levels;
  \item bagging introduces a bias, but sampling $(1- 1/e) \approx 0.632$ of the data without replacement seems to be an
    effective strategy for removing the problem.
  \end{enumerate}
\end{itemize}

@@

Biau~{\it et al.}~\cite{Biau.2012} provide a proof that a random forest model will converge at a rate that depends on
the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
rate that depends on the size of $S$ and not the number of noise variables in the data set. If the size of $S$ is small
compared to the total number of variables, then the rate of convergence will be quicker. It is not clear what the size
of the set $S$ will be in a GWAs application. There are cases of small groups of SNPs with large effects and also of
conditions caused by the interaction of large groups of SNPs.

@@

However, within an RDD, spark ML stores each sample as a vector. Unlike RDDs, which can be partitioned and distributed
across multiple nodes, each vector must be present in its entirety on any node accessing it. This is no problem with
typical datasets, however as dimensionality increases, the vectors eventually reach a size where they can no longer fit
into a single node's memory.

@@

The current implementation of \cursedforest uses a ``Gini impurity'' criteria for splitting. Let $f_q$ be the fraction
of items labelled with value $q$ where $q = 1, \ldots, Q$ at a node. The Gini impurity is
$$
I_G(f) = \sum_{q = 1}^Q f_q ( 1 - f_q ), 
$$
which is a minimum when all observations at the node are in the same class.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This section talks about the datasets
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Datasets}
\subsubsection{1000 Genomes Project}
We obtained the 1000 Genomes Project data as VCF files from their FTP site.  Each VCF file contains the variants for
every individual across one chromosome.  We use the phase 3 dataset which contains 2,504 individuals with approximately
81 million variant positions (i.e. a matrix of $2,504 \times 81$ million).  For compatibility with the majority of the
algorithms we demonstrate, we convert the original variant representations to double representations.  For example, we
store the absence of a variant (0\textbar0) as 0, a heterozygous
variant (1\textbar0 or 0\textbar1) as 1, and a
homozygous variant (1\textbar1) as 2.  However, this is not the only possible encoding \cite{Goldstein.et.al.2011}. Note that from
the encodings listed in Table~\ref{table:encodings}, the additive encoding will present the most difficulty for a random
forest model.

@@

\subsection{Parameter settings}
We consider the parameter setting for the random forest algorithm. We use the R notation from the \randomforest
package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
the advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree{}  -- the number of trees.  The number of trees necessary for good performance grows with the number of
  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree is necessary to get stable estimates of variable
  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
  it is possible for a random forest model to have a poorer fit and still have an accurate ranking of variable
  importance;
\item \mtry{}  -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry may give
  better performance;
\item the size and complexity of the individual trees is controlled in \randomforest by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest\footnote{The Spark ML documentation \cite{Spark.2016} sets \texttt{maxDepth} to 4 in their
    classification example. This may give an estimate with a higher bias and, as such, may be a poor choice. See the
    discussion in \cite{Dietterich.2002} where it is shown that bagging may reduce bias as well as variance but gives a
    better result for low bias learners. \cite{Dietterich.2002} notes that if the bootstrap replicate approximation were
    correct (i.e.~if the bootstrap sample came from an identical distribution to the data), then bagging would reduce
    variance without changing bias.} In the TCGA example the RandomForests code produced trees with depths of 30+.
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 2: Feature Selection on real data 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ selects previously demonstrated bone-density variants}
We investigate the ability of using a random forest model we build on genomic data to identify variants related to a
specific phenotype.
In this case we aim to identify variants associated with Bone Mineral Density (BMD). We build a model on the BMD dataset
(see datasets) using a binary high or low BMD value as the label.
From this model, we output a list of the top 10,000 contributing variants, ranked by their Gini impurity. The variant with
the highest Gini impurity---and hence, importance---is ranked 1, and the variant with the lowest Gini impurity is ranked 10,000.
Then, using the 26 known and 6 novel BMD genes from Duncan~{\it et al.}, we compile a list of variants that are present in or near one of these
32 genes.

@@

We also frequently see highly-ranked variants in chromosome 22q11. Stagi~{\it et al.} observe that individuals with a microdeletion
in this region have a significant in bone mass\cite{Stagi2010}. In this same region, we also observe variants present in
PRODH (proline dehydrogenase). PRODH catalyzes the first step in proline degradation, and as proposed by Phang~{\it et al.}, proline
is made available by collagen degradation \cite{Phang2008}.

DCDC5 (Doublecortin Domain Containing 5) is one of the 26 known genes, however it is one of the genes that Duncan~{\it et al.}.
failed to replicate in their study. DCDC5 does appear in our list of genes, however the first occurrence of a variant in
or near this gene has a rank of 9,667 out of 10,000. Also, this gene is not highly expressed in bone, and any mechanisms
by which this gene might regulate BMD are unclear \cite{Thakker2012}. Although variants near DCDC5 are associated with
BMD \cite{Rivadeneira2009}, it is questionable as to whether the gene itself contributes to BMD.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 3: Classification selection
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest outperforms existing methods for classification problems}
In this section we compare \cursedforest's performance against other published methods for predicting ethnicity on the
1000 Genomes Project dataset.

@@

\begin{figure}[tbhp] 
    \caption{\textbf{Runtime comparison on the 1000 genomes project}} 
    \label{figure:synth1}
      \includegraphics[totalheight=8cm]{./figs/1000genomesRuntime.pdf}
    \begin{flushleft} 
      The figure shows the runtime for the 5 subsets of the 1000 genomes project and the different methods investigated. $\bigotimes$ marks the last dataset the different method completed before running out of memory or running more than 24 hours. The dashed lines mark the trajectory for how ranger and Spark ML had performed given larger memory machines. 
    \end{flushleft}
\end{figure}

@@

We note several things here:
\begin{itemize}
\item the Donoho-Tanner phase transition arises in recovering the $\beta$ in data generated by a
  linear model. However, in a decision tree (random forest) there is no notion of estimating the $\beta$
\item A decision tree (random forest) is a heuristic search. It may recover a relationship in the space of the
  combinatorial search.
\end{itemize}

The existence of the  Donoho-Tanner phase transition is a salutary warning. There are likely to be limits, both
computational and logical, to the recovery of signals from noisy data. \cursedforest  is a contribution to addressing the
practical limits but the logical limits will still apply. However in the case of data that is both big and wide,
\cursedforest and other VariantSpark methods may provide a useful tool.
